Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-12 ebayes_rnaseq_heterosis.tex
added figures
start of manuscript

%  template.tex for Biometrics papers
%
%  This file provides a template for Biometrics authors.  Use this
%  template as the starting point for creating your manuscript document.
%  See the file biomsample.tex for an example of a full-blown manuscript.

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
%documentclass[useAMS]{biom}
%
%  If your system does not have the AMS fonts version 2.0 installed, then
%  remove the useAMS option.
%
%  useAMS allows you to obtain upright Greek characters.
%  e.g. \umu, \upi etc.  See the section on "Upright Greek characters" in
%  this guide for further information.
%
%  If you are using AMS 2.0 fonts, bold math letters/symbols are available
%  at a larger range of sizes for NFSS release 1 and 2 (using \boldmath or
%  preferably \bmath).
% 
%  Other options are described in the user guide. Here are a few:
% 
%  -  If you use Patrick Daly's natbib  to cross-reference your 
%     bibliography entries, use the usenatbib option
%
%  -  If you use \includegraphics (graphicx package) for importing graphics
%     into your figures, use the usegraphicx option
% 
%  If you wish to typeset the paper in Times font (if you do not have the
%  PostScript Type 1 Computer Modern fonts you will need to do this to get
%  smoother fonts in a PDF file) then uncomment the next line
%  \usepackage{Times}

@@

\title[Empirical Bayes analysis for detection of gene heterosis in RNAseq data]{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}

@@

\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
\volume{VV}
\pubyear{YYYY}
\artmonth{Month}
\doi{}

@@

\section{Hierarchical model}
\label{s:model}

\section{Empirical Bayes}
\label{s:inf}

@@

@@

Web Appendix A, referenced in Section~\ref{s:model}, is available with
this paper at the Biometrics website on Wiley Online
Library.\vspace*{-8pt}

>>>>

\documentclass[useAMS,usenatbib,referee]{biom}

\usepackage{amsmath}
\usepackage[figuresright]{rotating}

\graphicspath{{../temp/rna-seq-stan/fig-rm-stan_laplace/}}

@@

\title[eBayes for heterosis in RNAseq]{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}

@@

\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
\volume{VV}
\pubyear{YYYY}
\artmonth{Month}
\doi{x}

@@

\section{Empirical Bayes}
\subsection{Hierarchical model for RNAseq counts}
\label{s:model}

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,F$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$. As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$. Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$.

\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\delta_{vi}},\psi_g) 
\label{e:data}
\end{equation}

A hierarchical model is constructed for the sequencing depth, overdispersion, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\delta_{vi} \stackrel{ind}{\sim} N(0,\tau_\delta^2)$. The logarithm of the gene-specific over-dispersion parameters also follow a normal distribution $\log(\psi_g) \stackrel{ind}{\sim} N(\zeta_\psi,\tau_\psi^2)$. The variety-specific effects for each gene, $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$ are assumed to have a joint normal distribution $\mu_g \stackrel{ind}{\sim} N(\eta, \Sigma)$. We consider an \emph{independent} model where $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_V^2$ and a \emph{covariance} model where we estimate all parameters in $\Sigma$.

We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations, e.g. $\tau_\delta, \tau_\psi \stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters on the untruncated distribution. In the independent model, we also assume $Ca^+(0,3)$ on the diagonal elements while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+1$ and an identity scale matrix.

\subsection{Empirical Bayes}
\label{s:ebayes}

The parameters can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and $\theta_g = (\mu_g,\psi_g)$ and non-gene-specific parameters $\phi = (\delta_{11},\ldots,\delta_{Vn_v}, \zeta_\psi, \tau_\psi^2, \Sigma)$. We employ an optimization procedure to obtain the \emph{maximum a posterior} (MAP) estimate for $\phi$, i.e. $\hat{\phi}_{MAP}$ and then base gene-specific inference on the posterior conditional on this estimated hyperparameter. Fortunately, conditional on the hyperparameters the gene specific parameters are conditionally independent as in equation \eqref{e:condind}.

\begin{equation}
p(\theta|y,\hat{\phi}_{MAP}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\phi}_{MAP}) 
\label{e:condind}
\end{equation}

@@

\subsection{Coverage for our model}

\input{../temp/analyze-heterosis/FIGURES/coverage}

\subsection{Heterosis detection}

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.8\textwidth]{exampleROC0_1}}
\caption{Example ROC curves for the first simulation}
\label{f:roc}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.8\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves (AUC) for 3 different sample sizes (facets) for the methods under comparison for false positive rates below 0.1. Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
\end{figure}

@@
